Disclosing funding sources for open access publication fees: the Open APC

### 2016-05-30 article.Rmd
some minor improvements

The aim of this study was to examine how much German universities and research organisations spent on open access publication fees until 2015. Drawing on self-reported data that German institutions openly shared through the Open APC initiative, the analysis focused on the amount of institutional payments for publication fees, and compared these findings with those from related Austrian and British initiatives. We also asked how thoroughly self-reported articles were indexed in CrossRef, a DOI minting agency for scholarly literature, and analysed how institutional spending per articles was distributed over publishers and journal titles.

@@

Instead of re-using publisher and journal information contained in the Open APC data-set, we fetched bibliographic metadata for each article from CrossRef on May 19, 2016, on the basis of the reported DOI's. Although the Open APC initiative gathered metadata representing publishers and journals from CrossRef as well, this information was  retrieved at the time when the participating institutions submitted the cost information. 
However, CrossRef regularly updates metadata to represent ongoing mergers of publishing houses or name changes. A prominent example in this regard was the merger of the two large publishing houses Springer Business + Media and Nature Publishing Group announced on May 6, 2015, that now operates as Springer Nature. To reflect these changes in academic publishing, we decided to retrieve updated metadata from CrossRef for the whole Open APC data-set.

As a client, we used the R package rcrossref [@rcrossref], developed and maintained by the rOpenSci initiative[^6], to access CrossRef's REST API. We requested the XML-based format `application/vnd.crossref.unixsd+xml` in which full and abbreviated journal titles as well as the media types of ISSN's, the International Standard Serial Number used to identify journals, were distinguished. It also contained disambiguated publisher information, thus avoiding confusion about naming of publisher houses other studies were faced with when working with self-reported data [@woodward_2014]. In cases where no bibliographic information could be obtained, we used the Open APC values. Since CrossRef is not the only registration agency for DOI's, but also the agencies DataCite and Medra mint DOIs for scholarly work, we furthermore obtained the DOI agency for each article with the help of the rcrossref client.

Data collection in this study also involved obtaining cost data from related open data initiatives. To compare self-reported spending on open access publishing by Germany universities and research organisations with that of other initiatives, we reviewed the openly available spreadsheets from the Wellcome Trust [@wellcome_apc], the Austrian Science Fund FWF [@fwf_apc] and Jisc. For analysis, we obtained the amount of overall spending on publications fees, as well as on charges to publish in fully open access journal articles. In the case of FWF, we gathered the cost information from accompanying reports. We used the spreadsheet data to summarize  Wellcome Trust's and Jisc's spending, and converted the prices from GBP to Euro in accordance with the average Euro foreign exchange reference rates provided by the European Central Bank. Our comparison between the open data initiatives focussed on the last two years 2014 and 2015.Since Wellcome Trust's spending was reported for the periods 2013 - 2014 and 2014 - 2015, we referred to the exchange rates of the full two-years as we could not determine the actual dates of spending. We excluded Jisc's spending in 2014 because full currency information in the data-set was not available.

Data collection methods of the Open APC initiative and those of the others differed in some aspects. For instance, whereas the DOI was a mandatory element in the Open APC data template that the participating institution were required to report, publication identifier in the Wellcome Trust data might also be added through automated compliance checks. For this reason, and as our main focus is institutional funding for publication fees in Germany, we decided only to compare German spending with that of other initiatives, but not its size and distribution over publisher and journal titles, as well as the indexing coverage in CrossRef. However, we made use of the literature to discuss our findings in relation to the other initiatives.

@@

ggplot(my_apc, aes(euro)) + 
  geom_histogram(binwidth = 250, colour ="black", boundary = 0) + 
  scale_x_continuous(limits = c(0,8000)) +
  xlab("Publication fee in €") +
  ylab("Articles") +
  theme_bw() + science_theme

@@

@@

The data-set finally confirms the leading role of "mega-journals" in open access publishing, including the multidisciplinary PLOS ONE and the journals New Journal of Physics, Atmospheric Chemistry and Physics Discussions and Frontiers in Psychology, all of which publish contributions from all branches of their respective discipline. In general, an estimated 14 out of more than 10,000 journals registered in DOAJ in 2015 accounted for up to 15–20 % of all articles published in full open access journals [@Bj_rk_2015].

@@

In Germany, institutional spending on open access publication fees has increased over the years. With a share of …, the majority of open access articles German institutions reported to the Open APC initiative were published in fully open access journals. This presumably reflects the DFG funding policy which excludes the support of articles published in hybrid open access journals. The DFG has been financially supporting the implementation of central publication funds at more than 30 German universities since 2009. However, reviewing self-reported cost data from funders or countries that also support hybrid open access journals or open access books revealed a smaller proportions of payments in favor of articles in fully open access journals. Since open access publication fee spending is generally fragmented, we cannot answer whether German researchers avoid opting for open access when publishing in hybrid journals or use other budgets to pay publication fees required to make their work open access through these kind of journals.

In our study, CrossRef thoroughly indexed the analysed open access journal articles. Metadata representing publisher and journal titles could be gathered for 99 % of the self-reported article and successfully merged with the Open APC cost data. Using metadata from CrossRef, therefore, could reduce the extensive validation work of bibliographic information in self-reported data-sets on open access publication fees provided that the reporting of the DOI along with the expenses is made mandatory. Drawing on CrossRef would also increase the comparability of cost data for future negotiations with publishers on open access agreements, and the open access spending between open data initiatives that apply the same reporting standards, as its metadata represent the dynamic landscape of academic publishing in terms of ongoing mergers of publishing houses or name changes.

Another advantage of self-reported data-sets on the article-level is that they enable researchers and practitioners alike to study in which open access journals researchers from one institution actually publish, and to compare these findings with that of other universities or research organizations. For instance, our study revealed that the size of publication fee spending differed among the institutions with the  Max Planck Society accounting for almost 39 % of the overall articles. Many universities and research organization reported remarkably lower number of supported open access articles to the Open APC initiative. Using self-reported data, therefore, contributes to the understanding about how much and to what extent spending on open access publishing varies on the institutional level. This is particularly relevant given the increasingly important role open access publishing plays in recent negotiations between German universities and research organizations forming consortia on the one side and publishers on the other about financing scholarly publishing in future [allianz].

Our study is limited in some respects. One is that we cannot assess whether publishers and journals granted publication fee discounts seeing that the Open APC initiative does not track this kind of information. However, the large price ranges of particular journals suggests that varying pricing schemes are in place. Adding to this complexity, it seems that some institutions only paid parts of the publication fee. Take for instance the journal Nature Communication. Charges reported in our sample ranged between 2000 €, the DFG price cap, and 4.403 €. Such co-payments that involve several budgets are a discussed strategy to sustain publication funds at German universities [Fournier]. In another case, one university included its charges for participating in the German SCOAP consortia and presumably divided the sum by the articles published in SCOAP journals. In this consortia managed by the German National Library of Science and Technology payments were not directly made on the article level. Instead, subscription costs between a participating library and a publisher were reconciled, and the reduction transferred to the consortia to finance publications in SCOAP journals.

It must also be noted that reporting to the Open APC initiative is voluntary. Therefore, not all institutions in Germany that provide central funding of publication fees contribute cost data to this initiative. In a qualitative survey that also asked why German institutions are reluctant to share their cost data through the Open APC initiative one institution feared that increase in transparency would allow publishers to adjust prices in their favor. Others pointed out that the workload to produce such a data-set could be too extensive [@deppe_2015]. As there no registry of institutional open access funds or similar support structures exists, we cannot assess how many German universities and research organizations do not share their spendings on open access publication fees.

Our analysis on how institutional spending per articles was distributed over publishers and journal titles shows that open access publishing is diverse and concentrated at the same time. While we were able to identify 139 individual publishing houses that were supported by the German universities and research organizations, the distribution is highly skewed. 92 % of open access publication fee spending went to ten publishers, confirming a general high concentration of few publishers in current academic publishing. However, our study could not confirm that publications in open access journals owned by traditional publishing houses account for most of the spending on publication fees. Rather, open access publishers such as Public Library of Science (PloS), Copernicus GmbH or MPDI AG rank higher in our study than in the analyses of cost data in the UK. The study finally confirms the importance of mega-journals that account for a large share of all articles in fully open access journals. 
Conclusion

>>>>

The aim of this study was to examine how much German universities and research organisations spent on open access publication fees until 2015. Drawing on self-reported data from the Open APC initiative, the analysis focused on the amount of institutional payments for publication fees, and compared these findings with those from related Austrian and UK initiatives. We also asked how thoroughly self-reported articles were indexed in CrossRef, a DOI minting agency for scholarly literature, and analysed how institutional spending per articles was distributed over publishers and journal titles.

@@

Instead of re-using publisher and journal information contained in the Open APC data-set, we fetched bibliographic metadata for each article from CrossRef on May 19, 2016, on the basis of the reported DOIs. Although the Open APC initiative gathered metadata representing publishers and journals from CrossRef as well, this information was retrieved at the time when the participating institutions submitted the cost information. The Open APC initiative tracked the date of data submission and the contributors for every data submission with Git, a version control system, increasingly used for enabling reproducible research [@Ram_2013], and made this information available via GitHub to be transparent over time. However, CrossRef regularly updated metadata to represent ongoing mergers of publishing houses or name changes. A prominent example in this regard was the merger of the two large publishing houses Springer Business + Media and Nature Publishing Group announced on May 6, 2015, that operated as Springer Nature at the time of our study. To reflect these dynamics in academic publishing, we decided to retrieve updated metadata from CrossRef for the whole Open APC data-set.

As a client, we used the R package rcrossref [@rcrossref], developed and maintained by the rOpenSci initiative[^6], to access CrossRef's REST API. We requested the XML-based format `application/vnd.crossref.unixsd+xml` in which full and abbreviated journal titles as well as the ISSN media types, the International Standard Serial Number used to identify journals, were distinguished. It also contained normalised publisher information, thus avoiding confusion about naming of publishing houses other studies were faced with when working with self-reported data [@woodward_2014]. In cases where no bibliographic information could be obtained, we used the Open APC values. Since CrossRef was not the only registration agency for DOIs, but also the agencies DataCite and Medra minted DOIs for scholarly work, we furthermore obtained the DOI agency for each article with the help of the rcrossref client.

Data collection in this study also involved obtaining cost data from related open data initiatives. To compare self-reported spending on open access journal articles by Germany universities and research organisations with that of other initiatives, we reviewed the openly available spreadsheets from the Wellcome Trust [@wellcome_apc], the Austrian Science Fund FWF [@fwf_apc] and Jisc. For analysis, we obtained the amount of overall spending on publications fees, as well as on charges to publish in fully open access journal articles. In the case of FWF, we gathered the cost information from accompanying reports. We used the spreadsheet data to summarize  Wellcome Trust's and Jisc's spending, and converted the prices from GBP to Euro in accordance with the average Euro foreign exchange reference rates provided by the European Central Bank. Our comparison between the open data initiatives focussed on the last two years 2014 and 2015. Because Wellcome Trust's spending was reported for the periods 2013 - 2014 and 2014 - 2015, we referred to the average exchange rates of the full two-year period as we could not determine the actual invoicing dates from the data. We excluded Jisc's spending in 2014 because full currency information in the data-set was not available.

Data collection methods of the Open APC initiative and those of the others differed in some aspects. For instance, whereas the DOI was a mandatory element in the Open APC data template that the participating institution were required to report, publication identifier in the Wellcome Trust data was also added through automated compliance checks. Our first screening of the data-sets revealed that some articles lacked a DOI. For this reason, and as our main focus is institutional funding for publication fees in Germany, we decided only to compare German spending with that of other initiatives, but not its size and distribution over publisher and journal titles, as well as the indexing coverage in CrossRef. However, we made use of the literature to discuss our findings in relation to these Austrian and UK initiatives.

@@

@@

ggplot(my_apc, aes(euro)) + 
  geom_histogram(binwidth = 250, colour ="black", boundary = 0) + 
  scale_x_continuous(limits = c(0,8000)) +
  xlab("Publication fee in Euro") +
  ylab("Articles") +
  theme_bw() + science_theme

@@

@@

In Germany, institutional spending on open access publication fees has increased over the years, confirming the general trend of publication fees in academic publishing [@Davis_2011; @Laakso_2012; @Pinfield_Review_2015]. With a share of 99%, the majority of open access articles German institutions reported to the Open APC initiative were published in fully open access journals. This presumably reflects the DFG funding policy which excludes the support of articles published in hybrid open access journals. The DFG has been financially supporting the implementation of central publication funds at more than 30 German universities since 2009. However, in reviewing self-reported cost data from funders or countries that also support hybrid open access journals or open access books, we revealed smaller proportions of payments in favour of articles in fully open access journals. Because open access publication fee spending is fragmented, we cannot answer whether German researchers avoided opting for open access when publishing in hybrid journals or used other budgets to pay publication fees required to make their work open access through these kind of journals.

In our study, CrossRef thoroughly indexed open access journal articles in the Open APC data-set. We could gather metadata representing publisher and journal titles for 99 % articles, and successfully merged these information with the Open APC cost data. Using metadata from CrossRef, therefore, reduces the extensive validation work of bibliographic information provided that the reporting of the DOI along with the expenses is made mandatory. Drawing on CrossRef would also increase the comparability of cost data for future negotiations with publishers on open access agreements, and the open access spending between open data initiatives that apply the same reporting standards, as its metadata represent the dynamic landscape of academic publishing in terms of ongoing mergers of publishing houses or name changes.

Another advantage of self-reported data-sets on the article-level to disclose spending on open access publication fees is that they enable researchers and practitioners alike to study in which open access journals researchers from one institution actually publish, and to compare these findings with that of other universities or research organizations. For instance, our study revealed that the size of publication fee spending differed among the institutions with the  Max Planck Society accounting for almost 39 % of the overall articles. Many universities and research organization reported remarkably lower number of supported open access articles to the Open APC initiative. Using self-reported data, therefore, contributes to the understanding about how much and to what extent spending on open access publishing varies on the institutional level. This is particularly relevant given the increasingly important role open access publishing plays in recent negotiations between German universities and research organizations forming consortia on the one side and publishers on the other about financing scholarly publishing in future [@allianz].

This study is limited in some respects. One is that we cannot assess whether publishers and journals granted publication fee discounts seeing that the Open APC initiative does not track this kind of information. However, the large price ranges of particular journals suggests that varying pricing schemes were in place. Adding to this complexity, it is likely that some institutions only paid parts of the publication fee. Take for instance the journal Nature Communication. Charges reported in our sample ranged between 2000 €, the DFG price cap, and 4.403 €. Such co-payments that involve several budgets were a proposed strategy to sustain publication funds at German universities [@Fournier_2013]. In another case, one university included its charges for participating in the German SCOAP consortia and presumably divided the sum by the articles published in SCOAP journals. In this national consortia, managed by the German National Library of Science and Technology, payments were not directly made per article. Instead, subscription costs between a participating library and a publisher were reconciled, and the reduction transferred to the consortia to finance publications in SCOAP journals.

It must also be noted that reporting to the Open APC initiative is voluntary. Therefore, not all institutions in Germany that provide central funding of publication fees contribute cost data to this initiative. In a qualitative survey, asking why German institutions are reluctant to share their cost data through the Open APC initiative, one institution feared that increase in transparency would allow publishers to adjust prices in their favour. Others pointed out that the workload to produce such a data-set could be too extensive [@deppe_2015]. As no registry of institutional open access funds or similar support structures exists, we cannot assess the number of non-participants in Germany.

Our analysis on how institutional spending per articles was distributed over publishers and journal titles shows that open access publishing is diverse and concentrated at the same time. While we were able to identify 139 individual publishing houses that were supported by the German universities and research organizations, the distribution is highly skewed. 92 % of open access publication fee spending went to ten publishers, confirming a general high concentration of few publishers in current academic publishing. However, our study could not confirm that publications in open access journals owned by traditional publishing houses account for most of the spending on publication fees. Rather, open access publishers such as Public Library of Science (PloS), Copernicus GmbH or MPDI AG rank higher in our study than in the analyses of cost data in the UK.

This study finally confirms the leading role of "mega-journals" in open access publishing, including the multidisciplinary PLOS ONE and the journals New Journal of Physics, Atmospheric Chemistry and Physics Discussions and Frontiers in Psychology, all of which publish contributions from all branches of their respective discipline. In general, an estimated 14 out of more than 10,000 journals registered in DOAJ in 2015 accounted for up to 15–20 % of all articles published in full open access journals [@Bj_rk_2015].
